---
title: "An introduction to the atena package"
author:
- name: Beatriz Calvo
  affiliation:
  - &id Dept. of Experimental and Health Sciences, Universitat Pompeu Fabra, Barcelona, Spain
- name: Robert Castelo
  affiliation: *id
  email: robert.castelo@upf.edu
package: "`r pkg_ver('atena')`"
abstract: >
  atena provides methods to quantify the expression of transposable elements within R and Bioconductor.
vignette: >
  %\VignetteIndexEntry{An introduction to the atena package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
bibliography: bibliography.bib
---

```{r setup, echo=FALSE}
options(width=80)
```

# What are transposable elements

Transposable elements (TEs) ...

# Current available methods for quantiying TE expression

Available methods for quantifying TE expression can be classified into ...

* Filtering methods such as ERVmap [@tokuyama2018ervmap].

* Reassignment methods such as Telescope [@bendall2019telescope].

# TE annotations

TE annotations can be found in ...

# Using atena to quantify TE expression

Quantifying TE expression with atena consists in the following two steps:

1. Build a parameter object for one of the available quantification methods.

2. Call the TE expression quantification method `qtex()` using the previously
   built parameter object.

## Build a parameter object for ERVmap

To build a parameter object for ERVmap ...

## Build a parameter object for Telescope

To build a parameter object for Telescope ...

## Quantify TE expression with `qtex()`

To quantify TE expression we should call the `qtex()` method as follows.

# Session information

```{r session_info, cache=FALSE}
sessionInfo()
```

# References
